Query Optimization for Dynamic Imputation

### 2017-05-15 paper/algorithm.tex
Add qualitative results from other imputation methods

@@

\subsubsection{Hot-deck}

Hot-deck imputation is a nonparametric method in which missing values are replaced with randomly selected complete values.
This method preserves the distribution of the complete values.
We use the same $P$ as for mean value imputation, but we modify $T$ to account for the fact that hot-deck is blocking in our implementation.

>>>>

We note here that for every imputation operator inserted into the query plan, a new
statistical model is instantiated and trained. Though it is tempting to try to pre-train
imputation models while the analyst begins to make queries, the exact rows of data, set
of complete attributes, and set of attributes to impute are unknown until runtime, making
this strategy infeasible. Thus, if applicable, $T$ should take into account both
imputation model training and application.

@@

\subsubsection{Hot deck}

Random hot deck imputation is a nonparametric method in which missing values in a column are
replaced with randomly selected complete values from the same column, preserving the
distribution of the complete values. We use the same $P$ as for mean value imputation, but
we modify $T$ to account for the fact that hot-deck is blocking in our implementation.

### 2017-05-15 paper/experiments.tex
Add qualitative results from other imputation methods

For each query, we evaluate \todo{update: three different configurations}: \ProjectName{} optimizing for quality ($\alpha=0$) and \ProjectName{} optimizing for performance ($\alpha=1$) and as a baseline, we fully impute the tables used by each query then run the query on the imputed tables, simulating the process that an analyst would follow to apply existing imputation techniques.

\Cref{fig:runtimes} shows a summary of the performance results. The quality-optimized
queries are an \todo{update/validate: order-of-magnitude faster than the baseline. We get another order-of-magnitude speedup when optimizing for performance}. This performance
differential means it is feasible
to explore multiple imputation operations (including more expensive operators) when using
\ProjectName{}, in contrast to the traditional approach of fully imputing the dataset.

\begin{figure}
\includegraphics[width=\columnwidth]{figures/running_times_combined_bar.png}
\caption{Runtimes for the queries in~\Cref{tbl:queries} using reference
    chained-equation decision trees algorithm. $\alpha=0$ (resp. $\alpha=1$) is
    \ProjectName{} run at the highest (resp. lowest) quality setting. Baseline
    is full table imputation followed by running the query. Each query and each
    value of $\alpha$ was run 200 times.}
\label{fig:runtimes}
\end{figure}

@@

\todo{Distinguish between underlying imputation methods}
For $\alpha=0$ and $\alpha=1$, we pair up tuples and compute SMAPE for all queries. That is,
we compute tuple-wise absolute percentage deviations within each iteration of a query, and
we report this value averaged over all iterations.  We
can see that optimizing for quality indeed reduces the SMAPE of query results.  In general, the SMAPE relative 
to the baseline is low---\todo{update: between \lowsmapealphazero{} and
\highsmapealphaoneexacs{} percent}---indicating that on-the-fly imputation produces results similar to the baseline. Notably, the results with largest error correspond
most closely to one existing approach of dropping all \nullv{}s.

@@

\todo{Should this be here?}
For every imputation operator inserted into the query plan, a new statistical model is instantiated and trained before being used for imputation.
Although it is tempting to further optimize query execution in \ProjectName{} by pre-training imputation models while the analyst begins to make queries,
this is not desirable because the specific rows of data, set of complete attributes, and set of attributes to impute are unknown until runtime.

@@

\todo{Update: Planning times for our benchmark queries are reasonable, with mean planning
times across the benchmark queries ranging from \planninglowtime{} ms to
\planninghightime{} ms, constituting \planninglowtimepct{} to \planninghightimepct{} of
total runtime, respectively.

In all cases, the optimizer returned a query
plan within \planningmaxtime{} ms.} The choice of $\alpha$ has no impact on planning times,
because query selection is done after collecting the Pareto frontier.

\review{
However, \ProjectName{}'s planning algorithm is still exponential in the number of joins in the plan, and the use of Pareto sets exacerbates this exponential growth.
We explored the limits of our planning algorithm by constructing a series of queries with increasing number of joins over the CDC tables.
Planning times (~\Cref{fig:planning-times}) for queries involving 1--5 joins---a practical value for  real world exploratory queries---are less than 1 second.
We extended \ProjectName{} to support an approximation of Pareto sets to improve planning times for queries with 6 or more joins. Approximate Pareto sets
displayed a reduction in planning time linear with respect to the reduction in size of the final frontier. For queries with 6--8 joins, approximate sets achieved a reduction
in planning time between 14\% and 23\%, on average. Users can increase the level of approximation to reduce planning times further.
}
%\todobox{Jose: I'm not a fan of this line, I would remove}{Furthermore, at that point, it may make more sense to specify a coherent model over the entire base table instead.}

>>>>

We evaluate \ProjectName{} optimizing for quality ($\alpha=0$), optimizing for performance
($\alpha=1$), and targeting a balance of each. As a baseline, we fully impute the tables
used by each query and then run the query on the imputed tables, simulating the process that
an analyst would follow to apply existing imputation techniques.

\Cref{fig:runtimes} shows a summary of the performance results. The quality-optimized
queries are an order-of-magnitude faster than the baseline. We get another order-of-magnitude speedup when optimizing for performance. This performance
differential means it is feasible
to explore multiple imputation operations (including more expensive operators) when using
\ProjectName{}, in contrast to the traditional approach of fully imputing the dataset.

\begin{figure}
\includegraphics[width=\columnwidth]{figures/running_times_combined_bar.png}
\caption{Runtimes for the queries in~\Cref{tbl:queries} using reference
    chained-equation decision trees algorithm. Using $\alpha=0$, \ProjectName{}
    is run at the highest quality setting, while with $\alpha=1$, performance
    is prioritized. Each query and each value of $\alpha$ was run 200 times.}
\label{fig:runtimes}
\end{figure}

@@

To calculate SMAPE for $\alpha=0$, $\alpha=0.5$, and $\alpha=1$, 
we compute tuple-wise absolute percentage deviations within each iteration of each query, and
average this value over all iterations. We
can see that optimizing for quality indeed reduces the SMAPE of query results.  In general, the SMAPE relative 
to the baseline is low --- between \lowsmapealphazero{} and
\highsmapealphaoneexacs{} percent --- indicating that on-the-fly imputation produces results similar to the baseline. Notably, the results with largest error correspond
most closely to one existing approach of dropping all \nullv{}s.

In \Cref{table:smape}, we report results using the reference chained-equation
decision trees model. We also experiment with mean-value imputation and hot
deck imputation, as outlined in \Cref{sec:imputation}. One approach to
mean-value imputation, to facilitate non-blocking operation within the iterator
model, is to estimate the column mean from base table histograms. In this case,
imputations are identical no matter where the operator is placed in the query
plan. Moreover, runtimes are essentially identical regardless of the value of
$\alpha$ because the time cost of the imputation is relatively low no matter
how many tuples need to be imputed. In hot deck imputation, runtimes are also
similar across $\alpha$, because the cost of buffering
tuples and sampling at random is negligible compared to the rest of query
execution. In terms of accuracy, relative errors using hot deck are close to
zero; since the baseline, in this case, is hot deck imputation on the base
table, only skew in the distributions arising from filters and joins would lead
to larger errors. In some cases, skew does arise, and errors are on the same
magnitude as in the decision trees case. Indeed, \ProjectName{} yields larger
benefit when using relatively sophisticated, higher quality imputation
algorithms as opposed to simpler strategies that would even be quick to
implement on the base table.

@@

@@

Planning times for our benchmark queries are reasonable, with mean planning
times across the benchmark queries ranging from \planninglowtime{} ms to
\planninghightime{} ms, and ranging from \planninglowtimepct{} to
\planninghightimepct{} as a share of total runtime.  In all cases, the
optimizer returned a query plan within \planningmaxtime{} ms. The choice of
$\alpha$ has no impact on planning times, because query selection is done after
collecting the Pareto frontier.

\review{
However, \ProjectName{}'s planning algorithm is still exponential in the number of joins in the plan, and the use of Pareto sets exacerbates this exponential growth.
We explored the limits of our planning algorithm by constructing a series of queries with increasing number of joins over the CDC tables.
Planning times for queries involving 1--5 joins---a practical value for  real world exploratory queries---are less than 1 second.
We extended \ProjectName{} to support an approximation of Pareto sets to improve planning times for queries with 6 or more joins. Approximate Pareto sets
displayed a reduction in planning time linear with respect to the reduction in size of the final frontier. For queries with 6--8 joins, approximate sets achieved a reduction
in planning time between 14\% and 23\%, on average. Users can increase the level of approximation to reduce planning times further.
}
%\todobox{Jose: I'm not a fan of this line, I would remove}{Furthermore, at that point, it may make more sense to specify a coherent model over the entire base table instead.}

### 2017-05-15 paper/relatedwork.tex
Add qualitative results from other imputation methods

Designers of data stream processing systems frequently confront missing values
and consider them carefully in query processing. Often, if sensor error is the
cause of missing values, values can be imputed with high confidence. \todo{What does this mean?} \textcite*{fernandez2009inter} use feedback punctuation to dynamically
reconfigure the query plan for state-dependent
optimizations as data continues to arrive. One of the applications of this framework is to avoid
expensive imputations as real-time conditions change. 
%We prove our
%algorithm and cost model produces an optimal set of plans to choose from,
%given the constraints of the search space.

>>>>

Designers of data stream processing systems frequently confront missing values
and consider them carefully in query processing. Often, if a physical process
like sensor error is the cause of missing values, values can be imputed with
high confidence. \textcite*{fernandez2009inter} use feedback punctuation to
dynamically
reconfigure the query plan for state-dependent
optimizations as data continues to arrive. One of the applications of this framework is to avoid
expensive imputations as real-time conditions change. 
%We prove our
%algorithm and cost model produces an optimal set of plans to choose from,
%given the constraints of the search space.
